---
title: "Blogging with R"
author: "Bernardo D'Auria -- bernardo.dauria@uc3m.es"
date: "Tuesday March 26th, 2019"
# output: html_document
output: 
 ioslides_presentation:
   logo: "uc3m-logo.png"
copyright: "@2019 Bernardo D'Auria"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Quick links  

slides in [htlm version](slides_html.html)

online link to [these slides](https://brdauria.github.io/codingclubuc3m_talk)

link to the slides' [GitHub repository](https://github.com/brdauria/codingclubuc3m_talk)

link to the example [GitHub repository](https://github.com/brdauria/my_blog)

## What is a Blog?

### Definition of blog

A blog (shortening of **weblog**) is an online journal or informational website displaying information in the reverse chronological order, with latest posts appearing first. 

It is a platform where a writer or even a group of writers share their views on an individual subject.


### Famous blog platforms: 

[WordPress.org](http://wordpress.org/), 
[Ghost](https://ghost.org/), 
[Drupal](http://www.drupal.org/), 
[Joomla](https://www.joomla.org/), 
[Wordpress.com](http://wordpress.com/), 
[Medium](http://medium.com/),
[Squarespace](https://www.squarespace.com/), 
[Weebly](https://www.weebly.com/), 
[Typepad](http://www.typepad.com/), 
[Blogger](https://www.blogger.com/), 
[Tumblr](http://tumblr.com/) and 
[Wix](https://www.wix.com/)

<sub  style="display: block; text-align: right">
  [source (2019)](https://makeawebsitehub.com/choose-right-blogging-platform/)
</sub>

## The requisites of our blog

Out aim is to crate a simple blog by the following collection of tools:

- [R](https://www.r-project.org/) + [R Markdown](https://rmarkdown.rstudio.com/lesson-1.html)

- [RStudio](https://www.rstudio.com/) + [blogdown](https://bookdown.org/yihui/blogdown/)

- [git](https://git-scm.com/downloads) + [GitHub](https://github.com/)

- [Jekyll](https://jekyllrb.com/) + [GitHub Pages](https://pages.github.com/)

This will allow us to:

- **do not** need a personal web server

- easily include **formulas** and **statistical plot**  
    in our blog posts.
    
## R Markdown

**R Markdown** is a file format to make dynamic documents in **R**. 

An R Markdown document is written in   
**markdown** (an easy-to-write plain text format)   
and contains chunks of **embedded R code**.

### Example:
<ul>
  ````markdown
    # Plotting a simple Histogram 
    
    We generate a number `n=50` of 
    **Normal** disributed random variables 
    and plot them as an histogram
    
    ```{r}`r ''`
     hist(rnorm(50))
    ```   
  ````
</ul>
    
## Plotting a simple Histogram 

We generate a number `n=50` of 
**Normal** disributed random variables 
and plot them as an histogram

```{r}
 hist(rnorm(50))
```  

## blogdown + RStudio

`blogdown` is a package in R, developed by *Yihui Xie* (also author of `knitr`), that facilitates to build blogs with **R Markdown** in RStudio.

It is mainly tailored for the static generator **Hugo**, but following the author of the package, we tweak it to make it work with the static generator **Jekyll**.


### Book published on 2017: 

["blogdown: Creating Websites with R Markdown"](https://www.crcpress.com/blogdown-Creating-Websites-with-R-Markdown/Xie-Hill-Thomas/p/book/9780815363729)  
      by Y. Xie, A. P. Hill, A. Thomas
      
it is also available [on-line ](https://bookdown.org/yihui/blogdown/)


## git + GitHub

- Create a repository, for example name it `my_blog`,  
   and **initialize** it with a `README.md`

- Go to `Settings -> GitHub Pages`
    - set **Source** to **master branch**
    - select a **Jekyll theme**, for example to **Minimal**
  
- Edit your repository in the **main** page  
   and set its **website link**
   
To manipulate locally repositories from your coputer,
you need to download and istall [git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git).

- [Linux](https://git-scm.com/download/linux)  
- [Mac](http://git-scm.com/download/mac")   
- [Windows](https://gitforwindows.org/)  


## Associate RStudio with Github

- Then enable version control in RStudio 
     by going to  
    `Tools -> Global Options... -> Git/SVN`

- There write the location of the `git` executable

- Then `Create RSA key...` 
    or select one if you already have them.

- Then open the **Terminal**
   by going to  
   `Tools -> Shell...`

- and enter your details for your commits

    ```bash
    git config --global user.email "my.email@university.edu"
    git config --global user.name "my.username"
    ```
    
## Create a new RStudio project

- `File -> New Project... -> Version Control -> Git`
- Type the `url` of your GitHub repository
- Select a name and a location for your RStudio project 


Then

- assure to have the package `blogdown` installed

  ```r
  ## Install from CRAN
  install.packages("blogdown")
  ## Or, install from GitHub
  if (!requireNamespace("devtools")) install.packages("devtools")
  devtools::install_github("brdauria/blogdown")
  ```

## Configure for Jekyll on GitHub Pages

- Set the file content `_config.yml` as follows

  ```yaml
  title: My first blog
  description: >- # this means to ignore newlines until "email:"
    "This blog has been creatged during the mini-course 
    'Blogging with R' by Bernardo D'Auria on Tue March 26th, 2019"
  email: my.email@university.edu

  # Build settings
  markdown: kramdown
  exclude: ['*.Rmd'] # this is important to use with RMarkdown
  theme: minima
  plugins:
    - jekyll-feed # to serve posts
  ```

- Create the folder `_posts` 
  
## Configure for Jekyll on GitHub Pages

- Create the file `Gemfile`

  ```yaml
  source "https://rubygems.org"
  
  # To upgrade, run `bundle update github-pages`.
  gem "github-pages", group: :jekyll_plugins
  
  group :jekyll_plugins do
    gem "jekyll-feed"
  end
  ```
  
- Add to `.gitignore`
  ```bash
  _site/
  .sass-cache/
  .jekyll-cache/
  .jekyll-metadata
  ```

## Add an `index.html` page

- Create the file `index.Rmd`  

  ````md
     ---
      layout: home
      title: "Welcome to my Blog"
     ---
      
      ```{r, echo=FALSE}`r ''`
      library(ggplot2)
      
      ## Plot
      g <- ggplot(mpg, aes(cty))
      g + geom_density(aes(fill=factor(cyl)), alpha=0.8) + 
          labs(title="Density plot", caption="Source: mpg", 
               x="City Mileage", fill="# Cylinders")
      ```
      
  ````
  
- Run `knitr::knit('index.Rmd')`

## Make your FIRST post

- Create the file `_posts/2019-03-01-HelloWorld.Rmd`  

  ````markdown
      ---
      title: "Hello World!"
      author: "Bernardo D'Auria"
      date: "2019-03-01"
      ---
      Let us sum `1+1` by **R**.
      
      ```{r}`r ''`
      1+1
      ```
      
  ````
- Run  

  ```r
    owd = setwd('_posts/'); 
    knitr::knit('2019-03-01-HelloWorld.Rmd'); setwd(owd);
  ```

## The blog works!

- `Commit` and `Push`

Now going to the url of the **GitHub Pages** of our repository,

the blog <font color='red'>WORKS!</font>

### To run it locally:

You need to have **Jekyll** installed

- run `bundle install`  
- run `bundle exec jekyll serve` 
- open in the browser the url `http://127.0.0.1:4000/`

## Jekyll

`Jekyll` is a simple, blog-aware, static site generator perfect for personal, project, or organization sites. 

Jekyll is the engine behind GitHub Pages, which you can use to host sites right from your GitHub repositories. 

- find [here](https://jekyllrb.com/docs/installation/)
  the instructions to install for different platforms

- it may require to download and install a more recent version of [Ruby language](https://www.ruby-lang.org/en/downloads/)

### Example in class: 

run `jekyll new . --force`
in the blog's root folder

## Make Jekyll the generator for blogdown

- create the hidden file [`.Rprofile`](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/.Rprofile) in the project directory  

   ```bash
   options(
      blogdown.generator = 'jekyll', blogdown.method = 'custom',
      blogdown.subdir = '_posts', servr.daemon = TRUE,
      blogdown.base.dir = 'assets/figures/',
      blogdown.base.url = '{{ "assets/figures/" | absolute_url }}'
    )
    # IMPORTANT: please leave this comment line!
   ```
  
## Make Jekyll the generator for blogdown

- create a new folder, name it `R`, containing two files:  
      - [`build.R`](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/R/build.R)  
      - [`build_one.R`](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/R/build_one.R)
      
  You can download them by clicking on them   
  or you can look for them in the `R` directory in
  the [brdauria/blogdown-jekyll](https://github.com/brdauria/blogdown-jekyll) repository
  
- make `sw/hd` link `config.yaml` to `_config.yml`

  ```bash
  ln -s _config.yml config.yaml # on MAC/Linux
  mklink config.yaml _config.yml # on Windows
  ```
  
## Make your SECOND post

- Select in the toolbar `Addins -> New Post`  
- Type the following informations
  
  ```yaml
  Title:    "Hello Jekyll!"
  Author:   "Bernardo D'Auria"
  Date:     2019-03-05
  Category: graphs
  Format:   R Markdown (.Rmd)
  ```
- Click on `Done` 

A new file named `2019-03-05-hello-jekyll.Rmd`   
will be added in the `_posts` directory  
and it will be opened for modifications.

## Make your SECOND post

- Modify the YAML header of the post as follows 

  ````yaml
  ---
  title: Hello Jekyll!
  author: "Bernardo D'Auria"
  date: '2019-03-05'
  slug: hello-jekyll
  categories:
    - graphs
  tags: []
  
  layout: post
  ---
  ````
  
## Make your SECOND post

- and the body as the following

  ````r
  This is my first plot that cointains an R graph
  
  ```{r}`r ''`
  library(ggplot2)
  ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length)) + geom_point()
  ```
  
  and an external picture
  
  ![colored cats](https://goo.gl/TRS6bL)
  
  ```
  ````

you can download the file here [2019-03-05-hello-jekyll.Rmd](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/_posts/2019-03-05-hello-jekyll.Rmd)

## Make your SECOND post

- `build` and `serve your blog  
  `Addins -> Serve Site` 
 
   and in the `Viewer Pane`  
   you should see your blog  
   
   WORKING with <font color='red'>PICTURES!</font>


- `commit` and `push`  
  
  and you should see your blog also in   
  
  WORKING in <font color='red'>INTERNET!</font>

## Needing some formulas

What we finally miss is to include in our posts some  
nice formulas

- like this $\sqrt{\frac{9}{4}} = \frac{3}{2}$ 

- and this

  $$ \int_0^{\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2} $$
 
So let us prepare our blog and make our **third post**!
  
## Jekyll directory structure
<style> 
  .jekyllTree { margin-left: 2em; }
  .jekyllTree * { margin: 0px; padding: 0px; line-height: 1em; font-size: 0.8em; } 
</style>
<div class='jekyllTree'>  
  ```
  .
  ├── _config.yml
  ├── _data
  |   └── members.yml
  ├── _drafts
  |   ├── begin-with-the-crazy-ideas.md
  |   └── on-simplicity-in-technology.md
  ├── _includes
  |   ├── footer.html
  |   └── header.html
  ├── _layouts
  |   ├── default.html
  |   └── post.html
  ├── _posts
  |   ├── 2007-10-29-post-2.md
  |   └── 2009-04-26-post-1.md
  ├── _sass
  |   ├── _base.scss
  |   └── _layout.scss
  ├── _site # generated by jekyll and containing the website
  ├── .jekyll-metadata
  ├── Gemfile
  ├── Gemfile.lock
  └── index.html # can also be an 'index.md' with valid front matter
  ```  
</div>

## The theme `minima` directory structure

- run the command

  ```r
  system2('bundle', 'show minima')
  ```

  to get the `minima` gem location.
  
- then 
  ```r
  list.files('/usr/local/lib/ruby/gems/2.5.0/gems/minima-2.5.0')
  ```

## The theme `minima` directory structure
<div class='jekyllTree'>  
  ```
  /usr/local/lib/ruby/gems/2.5.0/gems/minima-2.5.0
  ├── _includes
  |   ├── disqus_comments.html
  |   ├── footer.html
  |   ├── google-analytics.html
  |   ├── head.html
  |   ├── header.html
  |   ├── icon-github.html
  |   ├── icon-github.svg
  |   ├── icon-twitter.html
  |   ├── icon-twitter.svg
  |   └── social.html
  ├── _layouts
  |   ├── default.html 
  |   ├── home.html
  |   ├── page.html
  |   └── post.html
  ├── _sass
  |   ├── minima
  |   |   └── ...
  |   └── minima.scss
  ├── assets
  |   ├── main.scss
  |   └── minima-social-icons.svg
  ├── LICENSE.txt
  └── README.md
  ```  
</div>

## Customize the blog to allow <font color='red'>formulas</font>

- in the folder `_include` create [`mathjax.html`](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/_includes/mathjax.html)

  ```html
  {% if layout.mathjax or page.mathjax %}
  <!-- https://docs.mathjax.org/en/
              latest/configuration.html#local-config-files -->
  <script type="text/javascript"
          src="https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.4/
               latest.js?config=TeX-AMS-MML_HTMLorMML,
               {{ 'assets/js/MathJaxLocal.js' | absolute_url }}">
  </script>
  {% endif %}
  ```

  It contains **Liquid templating** syntax!
  See [here](https://jekyllrb.com/docs/liquid/) for more!
  
- in the folder `assets/js` create [`MathJaxLocal.js`](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/assets/js/MathJaxLocal.js)
  
  ```
  # TOO BIG to include here. Please click to download!
  ```
  
## Customize the blog to allow <font color='red'>formulas</font>

- in the folder `_layout` create [`mpost.html`](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/_layout/mpost.html)

  ```html
  ---
  layout: post
  mathjax: true
  ---
  
  {% include mathjax.html %}
  
  {{ content }}
  ```
  
## Make your THIRD post

- Select in the toolbar `Addins -> New Post`  
- Type the following informations
  
  ```yaml
  Title:    "Hello Math!"
  Author:   "Bernardo D'Auria"
  Date:     2019-03-15
  Category: math
  Format:   R Markdown (.Rmd)
  ```
- Click on `Done` 

A new file named `2019-03-15-hello-math.Rmd`   
will be added in the `_posts` directory  
and it will be opened for modifications.

## Make your THIRD post

- Modify the YAML header of the post as follows 

  ````yaml
  ---
  title: Hello Math!
  author: "Bernardo D'Auria"
  date: '2019-03-15'
  slug: hello-math
  categories:
    - math
  tags: []
  
  layout: mpost
  ---
  ````
  
## Make your THIRD post

- and the body as the following

  ```markdown
  This is my first plot that cointains  
  nice formulas

  - like this 
  \\( \sqrt{\frac{9}{4}} = \frac{3}{2} \\) 
  
  - and this
  
    $$ \int_0^{\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2} $$
  ```

you can download the file here [2019-03-15-hello-jekyll.md](https://raw.githubusercontent.com/brdauria/codingclubuc3m_talk/master/docs/files/_posts/2019-03-15-hello-jekyll.md)

- see immediatelt the preview in the **Viewer Pane**  

- `commit` and `push` to check it in Internet

## Working with Themes

**Jekyll** is very powerful for   

- its modularity  

- its themes  

However this has a _cost_:  

- **things may work slightly different from _theme_ to _theme_**
  
therefore there is some **extra work** to do!


We are going to work with: [Minimal Mistakes](https://mmistakes.github.io/minimal-mistakes/) theme

## Minimal Mistakes theme

- remove `index.md` and write `index.html` as follows:
  ````yaml
     ---
      layout: home
      author_profile: true
     ---
  ````

- add to `Gemfile`  
  ````yaml
     group :jekyll_plugins do
      ...
      gem "jekyll-include-cache"
    end
  ````
  
## Minimal Mistakes theme

- add to `_config.yml` this
  ````yaml
     github: [metadata] # important to use remote_theme
     minimal_mistakes_skin: dark
     remote_theme: "mmistakes/minimal-mistakes@4.15.2"
     paginate: 5
     plugins:
      ...
      - jekyll-include-cache
      - jekyll-paginate
  ````

## Minimal Mistakes theme

- add to `_config.yml` and this
  ````yaml
      # Front Matter Defaults:
      defaults:
        # _posts
        - scope:
            path: ""
            type: posts
          values:
            layout: single
            author_profile: true
            read_time: true
            comments: true
            share: true
            related: true
  ````

- then `Commit` and `Push`

## More resources

**Jekyll**

  - [docs](https://jekyllrb.com/docs/)  
  
  - [tutorials](https://jekyllrb.com/tutorials/home/)  
  
  - [themes](http://jekyllthemes.org/)

online books  

  - [R Mardown](https://bookdown.org/yihui/rmarkdown/): The Definitive Guide  
  
  - [blogdown](https://bookdown.org/yihui/blogdown/): Creating Websites with R Markdown

## Thanks

[**Coding Club**](https://codingclubuc3m.github.io) UC3M

[**tidyverse**](https://ggplot2.tidyverse.org) ecosystem

Slides created via [**R Mardown**](https://github.com/rstudio/rmarkdown) package
and [**ioslides**](https://github.com/iosscripts/iosslider)

Blog created with [**Jekyll**](https://jekyllrb.com/) 

Hosted in [**GitHub**](https://github.com/)

Using [*RStudio*](https://www.rstudio.com/) and [**blogdown**](https://bookdown.org/yihui/blogdown/) package  
   
plus some tweaks of the *Yihui Xie*`s [**blogdown-jekyll**](https://github.com/yihui/blogdown-jekyll) repository 

